---
title: R's Atomic Data Types and Vectorization
layout: default_with_disqus
author: Eric C. Anderson
output: bookdown::html_chapter
---

# R Data Types {#r-data-types}
Gotta put this in there.







## Basic Data "Modes" of R
There are four main "modes" of scalar data, in order from least to most general:

1. `logical` can take two values: `TRUE` and `FALSE`, which can be abbreviated, when you type them as `T` and `F`.
1. The `numeric` mode comes in two flavors: "integer" and "numeric" (real numbers).  Examples: 1, 3.14, 8.2, 10, etc.
1. `complex`: these are complex numbers of the form $a + bi$ where $a$ and $b$ are real numbers and $i=\sqrt{-1}$. Examples: `3.2+7.3i`,  `4+0i`
1. `character`: these take values that are often called "strings" in other languages.  Examples: "fred", "foo", "bar", "boing".
There is also a `raw` mode which refers to raw bytes of data, but we won't concern ourselves with that for now.

## Atomic Vectors
A fundamental data structure in R: a vector in which every element is of the same mode. Like
```{r}
x <- c(1,2,3,5,7)
x
```
Pretty basic stuff, until you start accidentally, or intentionally mixing modes.
```{r}
x <- c(1,2,3,5,7,"11")
x
```
The mode of everything is _coerced_ to the mode of the element with the most general mode, and this can really bite you in the rear if you don't watch out!

## Coercion Up One Step

* logical to numeric:
    * `TRUE` ==> `1`

    * `FALSE` ==> `0`

*  numeric to complex:
    * `6.4` ==> `6.4+0i`

    * `5` ==> `5+0i`

*  complex to character:
    * `6.4+0i` ==> `"6.4+0i"`


## Coercion Up Two Or More Steps
Note that the coercion "jumps over the intermediate steps"

* logical to complex
    * TRUE -> 1+0i

    * FALSE -> 0+0i

*  logical to character
    * TRUE -> "TRUE"

    * FALSE -> "FALSE"

*  numeric to character
    * 7 -> "7"

    * 3.1415 -> "3.1415"


## Coercion down one step
If the coercion doesn't make sense you end up with `NA` which is how R denotes missing data

* numeric to logical (0 -> FALSE, anything else -> TRUE)
    * 0 -> FALSE

    * 1 -> TRUE
    * 78.2 -> TRUE
    * 0.0001 -> TRUE
    * -563.3 -> TRUE

*  complex to numeric (discards complex part and warns about it!)
      * 3.4+0i -> 3.4

      * 5.6+7.6i -> 5.6 (+ a warning)

*  character to complex
      * "3.4+4i" -> 3.4+4i

      * "a" -> NA

## Coercion down more than one step
Important point: it doesn't _necessarily_ go through intermediate steps:

* complex to logical (0->FALSE, anything else -> TRUE)
    * 0+0i -> FALSE

    * 0+2i -> TRUE
    * 5+0i -> TRUE
    * 5+9i -> TRUE

* character to logical
    * "TRUE" -> TRUE

    * "FALSE" -> FALSE
    * "1" -> NA  # yikes! if it went through numeric you'd get something different!
    * "0" -> NA

*  character to numeric
    * "56.764" -> 56.764

    * "4+8i" -> 4 (with a warning that the complex part was dropped)
    * "fred" -> NA


## Functions For Explicit Coercion
There is a whole family for coercing objects between different modes (or different types) that take the form `as.something`:

* `as.logical(x)`
* `as.numeric(x)`
* `as.integer(x)` # note a mode, but we will come back to this eventually
* `as.complex(x)`
* `as.character(x)`


As expected, these are vectorized---they coerce every element of the vector to the desired mode.



## Missing Data and Special Values in R
We saw `NA` up above.  That means "Not Available" and it denotes missing data.    

There are also two more interesting values:

1. `Inf` (-Inf) means \(\infty\) (or \(-\infty\)) and arises from things like: 1/0 or log(0).
1. `NaN` means "not a number" and it arises from situations where you can't evaluate something and it doesn't have an obvious limit.  Like 0/0 or Inf/-Inf  or 0*Inf.
If you wish to test whether something is NaN, or NA you have: `is.na(x)` and `is.nan(x)` which return logical vectors.



## Modes of Missing Data
Here is something to be aware of: missing values, like non-missing values, carry around their mode. Try this:
```{r}
x <- c(1, 2, NA, 4, "5")
x
x[3]  # this extracts the third element of x
c(10,20,30,x[3])
c(10, 20, 30, NA)  # this is a "fresh" NA, no coercion
```


## Most Operators are Vectorized
This is _incredibly important_!  All the mathematical operators, like `+`, `-`, `*`, and the logical operators, like  `&` (AND), `|` (OR), and the comparison operators, like `<` and `>` are hungry to operate _element-wise_ on every _element_ of a vector.  Example:
```{r}
fish.lengths <- c(121, 95, 87, 142)
fish.weights <- c(1011, 505, 702, 900)
fish.fatness <- fish.weights / fish.lengths
fish.fatness
```
