---
title: Making Maps With R
author: "Eric C. Anderson"
output:
  html_document:
    toc: yes
  bookdown::html_chapter:
    toc: no
layout: default_with_disqus
---



# Making Maps with R {#map-making-in-R} 

```{r, include = FALSE}
library(knitr)
opts_chunk$set(fig.width=10,
               fig.height=7,
               out.width = "600px",
               out.height = "420px",
               fig.path = "lecture_figs/making-maps-")
```

## Intro {#map-making-intro}

For a long time, R has had a relatively simple mechanism, via the `maps` package, for making simple outlines
of maps and plotting lat-long points and paths on them.

More recently, with the advent of packages like `sp`, `rgdal`, and `rgeos`, R has been acquiring much of the
functionality of traditional GIS packages (like ArcGIS, etc).  This is an exciting development, but not
always easily accessible for the beginner, as it requires installation of specialized external libraries
(that may, on some platforms, not be straightforward) and considerable familiarity with GIS concepts.

More recently, a third approach to convenient mapping, using `ggmap` has been developed that allows the tiling of 
detailed base maps from Google Earth or Open Street Maps, upon which spatial data may be plotted.
Today, we are going to focus on mapping using base maps from R's tried and true `maps` package and also using the
`ggmap` package.  We won't cover the more advanced GIS-related topics nor using `rgdal`, or `sp` to plot
maps with different projections, etc.  However, we will talk briefly about the somewhat more simplified
approach using the `mapproj` package.

As in our previous explorations in this course, when it comes to plotting, we are going to completely
skip over R's base graphics system and head directly to Hadley Wickham's `ggplot2` package.  Hadley has
included a few functions that make it relatively easy to interact with the data in R's `maps` package, and
of course, once a map layer is laid down, you have all the power of ggplot at your fingertips to overlay
whatever you may want to over the map.  `ggmap` is a package that goes out to different map servers and
grabs base maps to plot things on, then it sets up the coordinate system and writes it out as the base layer
for further ggplotting.  It is pretty sweet, but does not support different projections.

### Today's Goals

1. Introduce readers to the map outlines available in the `maps` package
    + Show how to convert those data into data frames that `ggplot2` can deal with
    + Discuss some `ggplot2` related issues about plotting things.
2. Look at how one might do a small number of different projections using the `mapproj` package.
3. Use `ggmap` to make some pretty decent looking maps

I feel that the above three topics should cover a large part of what people will need for making
useful maps of field sites, or sampling locations, or fishing track lines, etc. 

For today we will be skipping how to read in traditional GIS "shapefiles" so as to minimize
the number of packages that need installation, but keep in mind that it isn't too hard to do that
in R, too.



### Prerequisites
You are going to need to install a few extra packages to follow along with this lecture.
```{r, eval=FALSE}
# these are packages you will need, but probably already have.
# Don't bother installing if you already have them
install.packages(c("ggplot2", "devtools"))

# some standard map packages.
install.packages(c("maps", "mapdata"))

# the github version of ggmap, which recently pulled in a small fix I had
# for a bug 
devtools::install_github("dkahle/ggmap")
```




### Load up a few of the libraries we will use

```{r}
library(ggplot2)
library(ggmap)
library(maps)
library(mapdata)
```


## Plotting maps-package maps with ggplot

### The main players:

* The `maps` package contains a lot of outlines of continents, countries, states, and counties that have
been with R for a long time.  
* The `mapdata` package contains a few more, higher-resolution outlines.
* The `maps` package comes with a plotting function, but, we will opt to use `ggplot2` to plot the 
maps in the `maps` package.  
* Recall that `ggplot2` operates on data frames.  Therefore we need some way to
translate the `maps` data into a data frame format the `ggplot` can use.

### Maps in the maps package

* Package `maps` provides lots of different map outlines and points for cities, etc.  
* Some examples: `usa`, `nz`, `state`, `world`, etc.



### Makin' data frames from map outlines

* `ggplot2` provides the `map_data()` function.
    + Think of it as a function that turns a series of points along an outline into a data frame
    of those points.
    + Syntax:  `map_data("name")` where "name" is a quoted string of the name of a map in the `maps` or `mapdata`
    package
* Here we get a USA map from `maps`:
    ```{r}
    usa <- map_data("usa")

    dim(usa)
    
    head(usa)
    
    tail(usa)
    ```
* Here is the high-res world map centered on the Pacific Ocean from `mapdata`
    ```{r}
    w2hr <- map_data("world2Hires")

    dim(w2hr)

    head(w2hr)

    tail(w2hr)
    ```

### The structure of those data frames
These are pretty straightforward:

* `long` is longitude.  Things to the west of the prime meridian are negative.
* `lat` is latitude.
* `order`. This just shows in which order `ggplot` should "connect the dots"
* `region` and `subregion` tell what region or subregion a set of points surrounds.
* `group`.  This is _very important_!  `ggplot2`'s functions can take a group argument which 
controls (amongst other things) whether adjacent points should be connected by lines.  If they are
in the same group, then they get connected, but if they are in different groups then they don't.
    + Essentially, having to points in different groups means that `ggplot` "lifts the pen" when going between
    them.
    

### Plot the USA map

* Maps in this format can be plotted with the polygon geom.  i.e. using `geom_polygon()`.
* `geom_polygon()` drawn lines between points and "closes them up" (i.e. draws a line from the last
point back to the first point)
* You have to map the `group` aesthetic to the `group` column
* Of course, `x = long` and `y = lat` are the other aesthetics.

#### Simple black map
By default, `geom_polygon()` draws with no line color, but with a black fill:
```{r}
usa <- map_data("usa") # we already did this, but we can do it again
ggplot() + geom_polygon(data = usa, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3)
```

#### What is this coord_fixed()?

* This is very important when drawing maps.
* It fixes the relationship between one unit in the $y$ direction and one unit in the $x$ direction.
* Then, even if you change the outer dimensions of the plot (i.e. by changing the window size or the size
of the pdf file you are saving it to (in `ggsave` for example)), the _aspect ratio_ remains unchanged.
* In the above case, I decided that if every $y$ unit was 1.3 times longer than an $x$ unit, then  the
plot came out looking good.
    + A different value might be needed closer to the poles.
    
#### Mess with line and fill colors

* Here is no fill, with a red line.  Remember, fixed value of aesthetics go _outside_ the `aes` function.
    ```{r}
    ggplot() + 
      geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = NA, color = "red") + 
      coord_fixed(1.3)
    ```
    
* Here is violet fill, with a blue line.
    ```{r}
    gg1 <- ggplot() + 
      geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = "violet", color = "blue") + 
      coord_fixed(1.3)
    gg1
    ```
    
#### Adding points to the map

* Let's add black and yellow points at our lab and at the NWFSC in Seattle.
    ```{r}
    labs <- data.frame(
      long = c(-122.064873, -122.306417),
      lat = c(36.951968, 47.644855),
      names = c("SWFSC-FED", "NWFSC"),
      stringsAsFactors = FALSE
      )  

    gg1 + 
      geom_point(data = labs, aes(x = long, y = lat), color = "black", size = 5) +
      geom_point(data = labs, aes(x = long, y = lat), color = "yellow", size = 4)
    ```

#### See how important the group aesthetic is

Here we plot that map without using the group aesthetic:
```{r}
ggplot() + 
      geom_polygon(data = usa, aes(x=long, y = lat), fill = "violet", color = "blue") + 
      geom_point(data = labs, aes(x = long, y = lat), color = "black", size = 5) +
      geom_point(data = labs, aes(x = long, y = lat), color = "yellow", size = 4) +
      coord_fixed(1.3)
```

That is no bueno!  The lines are connecting points that should not be connected!








### State maps
We can also get a data frame of polygons that tell us above state boundaries:
```{r}
states <- map_data("state")
dim(states)

head(states)

tail(states)
```

#### Plot all the states, all colored a little differently

This is just like it is above, but we can map fill to `region` and 
make sure the the lines of state borders are white.
```{r}
ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)  # do this to leave off the color legend
```

Boom! That is easy.

#### Plot just a subset of states in the contiguous 48:

* Read about the `subset` command.  It provides another way of 
subsetting data frames (sort of like using the `[ ]` operator with
a logical vector).
* We can use it to grab just CA, OR, and WA:
    ```{r}
    west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
    
    ggplot(data = west_coast) + 
      geom_polygon(aes(x = long, y = lat), fill = "palegreen", color = "black") 
    ```

#### Man that is ugly!! 

* I am just keeping people on their toes. What have we forgotten here?
    + `group`
    + `coord_fixed()`
* Let's put those back in there:
    ```{r}
    ggplot(data = west_coast) + 
      geom_polygon(aes(x = long, y = lat, group = group), fill = "palegreen", color = "black") + 
      coord_fixed(1.3)
    ```

Phew! That is a little better!


Now, can we add some county lines on there?
```{r}
counties <- map_data("county")
ca_county <- subset(counties, region == "california")
wc2 + geom_polygon(data = ca_county, aes(x = long, y = lat, group = group), color = "black", fill = NA )
```
That is rad.  Now I can go to wikipedia or http://www.california-demographics.com/counties_by_population
and grab population data for each county.

I can also get the square miles from wikipedia so I can do full on population density.




### zoom in?
Note that the scale of these maps kinda sucks!

```{r}
wc3 <- wc2 + xlim(-123, -121.5) + ylim(36, 38)
wc3
```

That is an epic fail because it clips the data and tries to complete the polygon with that ridiculous line

Better to chop it off using `coord_cartesian()`:
```{r}
wc4 <- wc2 + coord_cartesian(xlim = c(-123, -121.5),  ylim = c(36, 38))
wc4
```
    
