---
title: Natural Earth Data Maps With R
author: "Eric C. Anderson"
output:
  html_document:
    toc: yes
  bookdown::html_chapter:
    toc: no
layout: default_with_disqus
---



# Using Natural Earth Data with ggplot {#nat-earth-data-maps} 

```{r, include = FALSE}
library(knitr)
opts_chunk$set(fig.width=10,
               fig.height=7,
               out.width = "600px",
               out.height = "420px",
               fig.path = "lecture_figs/nat-earth-ggplot")
```

## Intro 

Found a blog post at http://downwithtime.wordpress.com/2013/12/04/naturalearthdata-and-r-in-ggplot2/
that had some stuff about it.

Going to try to follow it.

### Install rgdal

This apparently needs gdal.  So try this:
```
2014-12-07 21:33 /~/--% brew install gdal
Error: You must `brew link libpng' before gdal can be installed
2014-12-07 21:33 /~/--% brew link libpng
Linking /usr/local/Cellar/libpng/1.6.10... 18 symlinks created
2014-12-07 21:33 /~/--% brew install gdal
==> Installing dependencies for gdal: jpeg, giflib, libtiff, lzlib, proj,
==> Installing gdal dependency: jpeg
==> Downloading https://downloads.sf.net/project/machomebrew/Bottles/jpeg-8d.mav
######################################################################## 100.0%
[...]
```
That installs a boatload of dependencies.  However, the build failed because it couldn't find libjpeg.

So, I tried brew intalling it and got this:
```
2014-12-07 21:35 /~/--% brew link libjpeg
Linking /usr/local/Cellar/jpeg/8d... 
Error: Could not symlink bin/wrjpgcom
Target /usr/local/bin/wrjpgcom
already exists. You may want to remove it:
  rm /usr/local/bin/wrjpgcom

To force the link and overwrite all conflicting files:
  brew link --overwrite jpeg

To list all files that would be deleted:
  brew link --overwrite --dry-run jpeg
2014-12-07 21:36 /~/--%  brew link --overwrite --dry-run jpeg
Would remove:
/usr/local/bin/wrjpgcom
/usr/local/bin/rdjpgcom
/usr/local/bin/jpegtran
/usr/local/bin/djpeg
/usr/local/bin/cjpeg
/usr/local/include/jpeglib.h
/usr/local/include/jmorecfg.h
/usr/local/include/jerror.h
/usr/local/include/jconfig.h
/usr/local/lib/libjpeg.a
```
So I tried it
```
brew link --overwrite jpeg
```
Then a quick
```{r, eval=FALSE}
install.packages("rgdal", type = "source")
```
did it for me.

Great.

### Getting on with it:

#### Load libraries
```{r}
library(raster)
library(rgdal)
library(ggplot2)
library(reshape2)
library(plyr)
```

#### Download Natural Earth Data
I grabbed the NE2 stuff at 1:50 with water from a download button at: 
http://www.naturalearthdata.com/downloads/50m-raster-data/50m-natural-earth-2/  I downloaded
the "small size" which was 83 Mb.

I just got the rasters for now and put them into `/Users/eriq/Maps/NE2_50M_SR_W/`

#### Just do the base raster layer

```{r}
# load raster stack
nat.earth <- stack('~/Maps/NE2_50M_SR_W/NE2_50M_SR_W.tif')

# I am sure this could be done faster with dplyr, now
# but this is how the guy did it:

#  I have a domain I'm interested in, but there's no reason you can't define something else:
quick.subset <- function(x, longlat){
 
# longlat should be a vector of four values: c(xmin, xmax, ymin, ymax)
  x@data$id <- rownames(x@data)
 
  x.f = fortify(x, region="id")
  x.join = join(x.f, x@data, by="id")
 
  x.subset <- subset(x.join, x.join$long > longlat[1] & x.join$long < longlat[2] &
                           x.join$lat > longlat[3] & x.join$lat < longlat[4])
 
  x.subset
}

# here is the part we want to focus on:
domain <- c(-98.6, -66.1, 36.5, 49.7)

# here he crops the nat earth base layer
nat.crop <- crop(nat.earth, y=extent(domain))

# this must make a data frame out of the raster:
rast.table <- data.frame(xyFromCell(nat.crop, 1:ncell(nat.crop)),
                         getValues(nat.crop/255))


# this adds color values in there as a single column, using the
# channels that are named NE2_50M_SR_W.{1,2,3}
rast.table$rgb <- with(rast.table, rgb(NE2_50M_SR_W.1,
                                       NE2_50M_SR_W.2,
                                       NE2_50M_SR_W.3,
                                      1))

```

Now, we should be able to plot the base layer:
```{r}
g1 <- ggplot(data = rast.table, aes(x = x, y = y)) +
  geom_tile(fill = rast.table$rgb)
```

That seems to take a very long time. But mostly that is the time it takes for the
thing to get rendered to the screen. But I wonder if `geom_raster` would be faster. It is
supposed to more efficient...
```{r}
g2 <- ggplot(data = rast.table, aes(x = x, y = y)) +
  geom_raster(fill = rast.table$rgb)
```
That certainly returns from the function quickly.  It might be a little quicker.

I wonder how long it takes to make a png or a jpeg?
```{r}
ggsave(filename = "ne.png", g2)
```

```{r}
ggsave(filename = "ne.jpg", g2)
```

#### Now, do another part of the world:

I am going to do a bit of BC:
```{r}
# note extent has to be given as min/max for long then lat
domain <- c(-128.673958, -122.016243, 48.126320, 51.114388)
nat.crop <- crop(nat.earth, y=extent(domain))
rast.table <- data.frame(xyFromCell(nat.crop, 1:ncell(nat.crop)),
                         getValues(nat.crop/255))
rast.table$rgb <- with(rast.table, rgb(NE2_50M_SR_W.1,
                                       NE2_50M_SR_W.2,
                                       NE2_50M_SR_W.3,
                                      1))
```

```{r}
bc1 <- ggplot(data = rast.table, aes(x = x, y = y)) +
  geom_raster(fill = rast.table$rgb)
bc1
```

Wow! That is totally unsuitable!!  So, this is a very coarse resolution map.

but the 1:10 couldn't be that much better.  Did I downscale stuff when I made our
big map.  


#### Try just the coastline

```{r}
# note that it apparently does not do tilde expansion on the path
ne_coast <- readOGR(dsn = "/Users/eriq/Maps/NaturalEarth", layer = "ne_50m_coastline")

coast.subset <- quick.subset(ne_coast, domain)
```

The try to plot it:
```{r}
ggplot() + geom_path(data=coast.subset, aes(x = long, y = lat, group = group), color = 'blue')
```

Boy! That is intended for very zoomed out things.  

What if I do the 1:10 verion?
```{r}
# note that it apparently does not do tilde expansion on the path
ne_coast <- readOGR(dsn = "/Users/eriq/Maps/ne_10m_coastline", layer = "ne_10m_coastline")

coast.subset <- quick.subset(ne_coast, domain)
```
And here we plot it:
```{r}
ggplot() + geom_path(data=coast.subset, aes(x = long, y = lat, group = group), color = 'blue')
```


## What if you use ggmap
Check this out:
```{r}
library(ggmap)
vi <- get_map("Vancouver Island", zoom = 7)
ggmap(vi)
```
And if you save that as a jpg or png you can zoom in even further. I don't think that 
Natural Earth data can hold a candle to that.


