---
title: R and Rstudio Orientation
layout: default_with_disqus
author: Eric C. Anderson
output: bookdown::html_chapter
---

# R and Rstudio Orientation {#rstudio-orientation}

Let's learn about:

1. The idea of RStudio "Projects"
1. The layout of the RStudio IDE
1. How the R console works


## Rstudio... 

* Is an IDE (integrated development environment) that _sits on top of_ R
and makes it easier to interact with R.
* Organizes your work into R in neatly-contained packages of work (typically data
and code) called "projects"
* Nothing mysterious about these---just collections of files stored together in
a single directory on your computer.

### Let's make a new project

Do This in RStudio:

1. Select File -> New Project 
1. Create project from:  Choose "New Directory"
    1. Choose "Empty Project"
    1. Give the directory name as `project1` 
    1. and put it somewhere in your home directory (browse to where you want to put it in the "create project as subdirectory of:")
    1. For now, unselect buttons to create a git repository or to use packrat.
    1. Hit `Create Project`

### Now Explore:

* Notice that RStudio has started R in your console in the project directory:
```{r, eval=FALSE}
getwd()  # this tells you the current directory that R is in

# here is what it is on my computer:
[1] "/Users/eriq/project1"
```

* Paste the following code into the console window and see what happens.  Do not worry if you
don't know what any of it means.
```{r, eval=FALSE}
# plot some linearly-related simulated data
set.seed(5) # set the random number seed
n <- 400
x_data <- rnorm(n, mean = 100, sd = 15)
y_data <- x_data + rnorm(n, mean=0, sd = 5)
plot(x_data, y_data)
fit <- lm(y_data ~ x_data)
abline(fit)

# don't know what the abline function does? Check out its help:
?abline
```

* Now we will take a tour of some of the windows
    + Environment
    + History
    + Plots
    + Files
    + Help
    + etc
    

## The R Console {#r-console}

### "Communicating" with R

* R is an _interpreted_ language, so it just reads through source code line by line and executes the instructions it finds.
* Contrast to _compiled_ language like C in which all the source code is parsed and checked for
consistency, and then binary computer instructions are made from that and optimized for speed, memory efficiency, etc.
* Because R goes line by line, we can work _interactively_ with it
* You can also write programs in a text editor and then send them to R, all at once, to be interpreted.
    + This is much more reproducible than interactive work
    + But interactive work is good for testing things, etc.



### The R Console
This is R's "command prompt" where you type _expressions_ and, when you hit RETURN, these expressions are _evaluated_
(so long as they are _complete_ expressions). Here is what it looks like:
```{r, eval=FALSE}
>
```
That little `>` symbol at the beginning of the line says "R is ready to begin a new expression." Now try entering an expression:
```{r}
13+19
```
It evaluated "13+19" and, by default, if "nothing is done" with the result of the evaluation, R just prints the result. We'll talk about the significance of the [1] later.


### Inserting Comments in Your Code

* Though it doesn't seem all that useful when working in interactive mode, you can write comments in your code.
* These let you point things out in your code, but not confuse the interpreter.
* Everything right of a `#` until the end of the line is ignored by the interpreter:
```{r}
13+19 # Look at me! I'm adding two primes
```



### Command Recall
Using your up arrow will recall the expressions that have been evaluated recently by the R interpreter.  This can be useful when working interactively.


### A Semicolon is as good as a RETURN
You can break a line into multiple expressions that get evaluated individually by separating them with semicolons:
```{r, eval=FALSE}
> 13+19; 12+2
[1] 32
[1] 14
```
When R gets to each semicolon, it evaluates the expression and then, because we aren't doing anything else with it, it prints the result.

### Incomplete Expressions Make R wait for you
If you press RETURN and R finds that the expression you typed is _incomplete_ then it taps its foot and waits for you to input the rest of the expression, to make it _complete_.  To show you that it is tapping its foot, R changes the prompt from `>` to `+`
```{r, eval=FALSE}
13+   # This expression is incomplete!  
+  19   # when I enter this, R has a complete expression and can evaluate it!
```
If you force R to evaluate an incomplete expression by giving it a `;` at the end of it, R will complain by giving an _error message_
```{r, eval=FALSE}
13+  # incomplete expression
+ ;    # I don't care R! Evaluate it anyway you lousy bum!
Error: unexpected ';' in:
"13+
;"
```
Note that this is useful if you don't know why your expression is incomplete and you want to start all over again anyway.


### R is more than a glorified calculator
Please tell me that R does more than just adding numbers together!! Please, please!

You're in luck.  Though your R expressions can be as simple as adding two numbers together, the power of R starts to be realized when you assign _values_ to _symbols_, because you then can evaluate expressions involving those symbols.  




## Symbols {#intro-symbols}

* A _symbol_ in R is a text string in your R code that refers to an _object_.  An _object_ is the name given to "anything that the R interpreter might know about."  One might say that "everything in R" is an object.
* In vernacular, some might refer to a symbol as a _variable_
* For R to recognize a symbol, that symbol must adhere to some standards:
    + Use only letters or numerals: (a-zA-Z0-9) as well as `.` and `_`
    + Must start with a letter or `.` 

What (if anything) is wrong with each of these:  `my.variable`, `your%variable`, `foo_bar`, `_bar_foo`, `.boing`, `Fish_lengths`,  `9_treatment_results` 



### Assigning Values
The special operators:  
  `<-`  and  `->`  
are used to _assign_ values to the objects referred to by symbols.  

* It is like a pointy arrow or a stick, so you can think of it is "stick this in the object"
```{r}
x <- 13    # leftwards assignment is typically used
19 -> y    # rightwards assignment is seldom used
z <- x + y   # assign result of the evaluated expression

# hey! it didn't print any result!
z  # just type the symbol to print it
```

* If you assign a result, there isn't anything left to print.  But if you just type a symbol, the value of the object it refers to is printed.  This is called "Auto Printing".



### R Assigns Values, not References
* In some languages, like C, there are facilities to assign a variable to "point at" a specific area in the computer memory where the value for a symbol is stored.
* That is _emphatically not_ what happens in R.  Observe:
```{r}
x <- 19
y <- x
y
x <- 13
x
y
```
If you assign the value of object1 to object2, the value of object2 will not change if you subsequently change the value of object1.  

### I Need More Than Scalars!!
Data almost always come as more than one number.  If you caught a lot of fish and recorded their lengths in mm it would _way-totally-super-uncool_ if you had to do this:
```{r}
fish1 <- 121; fish2 <- 95; fish3 <- 87
fish4 <- 142  # and so on and on, so help me God!
```
But R has you covered here.  R loves _vectors_! (one dimensional arrays), observe:
```{r}
fish <- c(121, 95, 87, 142)
fish
```


## A quick word on functions {#quick-intro-to-functions}

* In the previous slide we saw the `c` function, which is the first _function_ we have seen that actually looks like a function (technically `x+y` is `"+"(x,y)`, but let us not get into that now).
* In R, functions typically take the form of:  
  `function_name( function_arguments )`  
where `function_name` must follow the requirements for symbols and the `function_arguments` could be lots of different things, as long as they make sense for the function.
* You pass arguments (which are typically other symbols that have some values assigned to them) to a function and the function returns a result. 

### The c() function
In
```{r}
fish <- c(121, 95, 87, 142)
```
The arguments to `c` are `121`, `95`, `87`, and `142`.

* The arguments could be _symbols_, of course:
```{r}
fish1 <- 121; fish2 <- 95; fish3 <- 87; fish4 <- 142
catfish <- c(fish1, fish2, fish3, fish4)
catfish
```

* The `c` function simply _catenates_ its arguments together and returns the result as a _vector_. 
* R treats everything as a vector: that `[1]` in its output tells us that the next thing it prints is the first element of a vector!



