---
title: "Homework 3"
author: "Eric C. Anderson"
output: bookdown::html_chapter
layout: default_with_disqus
---


## How To Do This Homework
To do this homework, you have to: 

1. __Pull down the master branch of
the rep-res-course repository to get this assignment__:
    * _If you don't already have a clone of the rep-res-course repository_, clone the rep-res-course repository:
        1. Choose File->New Project
        1. When asked, choose “Create Project from Version Control”
        1. Then when asked choose "Git"
        1. For repository URL use this: https://github.com/eriqande/rep-res-course.git
        1. Let the project directory be named `rep-res-course`. Put it somewhere. You might consider having a dedicated directory for all your work in this course.
        1. Hit “Create Project”.
    * _If you already have a clone of the rep-res-course repository_
        1. Open up the RStudio project `rep-res-course.Rproj`
        1. If you are not on branch `master` in that project (check the upper
        right corner of the "Git" pane---if it does not say "master" in small letters you are not 
        on the master branch), then __commit any changes on whatever branch you are on__ (If you don't commit
        changes on the branch you are on, then they will get carried back
        to the master branch and could cause problems pulling.), and then checkout
        the master branch (this can be done by clicking on the branch name in the upper right of the Git pane
        and choosing `master` (__not__ `origin/master`)).    
        1. Once on the `master` branch click the "pull button at the top of the Git pane.
1. __Immediately checkout a new branch named ``r rr_branch_name``__:
    1. go to the “Tools” menu item in RStudio and within it choose "Shell...". This will pop up a command prompt window (Terminal window on a mac, GitBash Window or a Command Prompt Window on PC).
    1. In the command window, type this command:
        ```{r, echo=FALSE, comment=NA}
        cat(paste("git checkout -b", rr_branch_name))
        ```
        and hit RETURN.
1. __Within RStudio, open the file ``r rr_hw_file_name`` and add your answers__ 
    1. (see [How to write answers](#writing-answers) below)
    1. Note that you can knit the whole document to see your answers given.  You can also 
    evaluate all the R code in a chunk to get the objects in it loaded into your R global
    environment.  See the options uner the menu item `Code --> Run Region` and get familiar
    with the keyboard shortcuts.
1. __Commit your completed homework file, and push your__ ``r rr_branch_name`` __branch up
to your fork of Eric's__ `rep-res-course` __repository__.
    1. _If you have not already done so, create a fork of the_ `rep-res-course` _repository on
    your own GitHub account_:
        + Go to https://github.com/ and make sure you are signed in with
        your username. (If you aren't signed in it
        will ask you to do so.)
        + Go to the `rep-res-course` on GitHub: https://github.com/eriqande/rep-res-course
        and hit the __Fork__ button in the upper right.
    2. _If you have not already done so, tell git on your computer to link your_ ``r rr_branch_name``
    _branch with your forked version
    of the `rep-res-course` repository_.  Use Tools --> Shell... to get a command prompt, then 
    replace GITHUBNAME with your GitHub username in the following:
        ```{r, echo=FALSE, comment=NA}
        cat(paste("git remote add my_rr_fork https://github.com/GITHUBNAME/rep-res-course.git"))
        ```
    You shouldn't have to do this more than once this quarter.
    3. Push your answers up to your fork of the repository. Give this command in
    the "Tools --> Shell" command shell:
        ```{r, echo=FALSE, comment=NA}
        cat(paste("git push my_rr_fork", rr_branch_name))
        ```
    You might have to provide your GitHub username and password.
    4. Go back to your GitHub site and press the large green button asking you if you want
    to make a pull request.
    5. Sit back and enjoy a moment.  You're done!

### How to write answers {#writing-answers}

1. I have written an R package that facilitates answer submission and collation
(and subsquent analysis) of homework responses.  You will need that package.  To get
it, use the `devtools` package.  You will have to install the `devtools` package
only if you don't already have it. So, 
    + issue the following expression in the R console _at least once this quarter_:
        ```{r, eval=FALSE}
        install.packages("devtools")
        ```
    + and do this in the R console (only one time) when you start working on the homework:
        ```{r, eval=FALSE}
        devtools::install_github("eriqande/rrhw")
        ```
2. Find the R code blocks within ``r rr_hw_file_name`` that follow the sections that look like:
__`r rr_homework_name`: #3, "problem-name"__.  You won't actually see that text written in
the Rmd document because it is generated automatically.  However, you can find the R code
chunk that is named, in this case, `problem-name` by searching for that string in the Rmd
file.
3. Read the problem, evaluate the lines in the R code chunk and maybe look at some
objects, and read the questions in the comments and then put your answer inside the
curly braces within the `submit_answer` function.  For example, a question that 
looks like this:
    ```{r, eval=FALSE}
    # x is a vector constructed like this:
    x <- 1:10
    
    # return the values of x less than 4
    submit_answer({
    
    })
    ```
should be answered like this:
    ```{r, eval=FALSE}
    # x is a vector constructed like this:
    x <- 1:10
    
    # return the values of x less than 4
    submit_answer({
    x[x<4]
    })
    ```
4. Note that you can put multi-line expressions into the `submit_answer` function.  This is crucial 
when the question asks you to modify and object _and also_ return the modified object.  For example
if given this question:
    ```{r, eval=FALSE}
    # x is a vector constructed like this:
    x <- 1:10
    
    # change all the even values in x to -99 AND return x
    submit_answer({
      
    })
    ```
should be answered like this:
    ```{r, eval=FALSE}
    # x is a vector constructed like this:
    x <- 1:10
    
    # change all the even values in x to -99 AND return x
    submit_answer({
      x[c(F,T)] <- -99
      x
    })
    ```
    Notice how the last line is `x`, which returns its value to the `submit_answer` function.


