---
title: "Homework 3"
author: "Eric C. Anderson"
output: bookdown::html_chapter
layout: default_with_disqus
---

```{r pre-stuff, echo=FALSE}
library(knitr)
submit_answer <- function(x) {
  ret <- list()
  ret$expr <- substitute(x)
  ret$estr <- deparse(substitute(x))
  ret$val <- eval.parent(substitute(x))
  print(ret$val)
  assign("boogiewho", ret, envir = .GlobalEnv  )
}

# here are variables that need to get set
rr_homework_name <- "Homework_3"
rr_question_number <- 0


# here is a hook to run on any chunk that has the rr.question set
knit_hooks$set(rr.question = function(before, options, envir) {
  if (before) {
    ## code to be run before a chunk
    if(grepl("^unnamed-chunk", options$label)[1]) stop("You have to name the chunk if it is an rr.question chunk")
    assign("rr_question_chunk_name", options$label, envir = .GlobalEnv)
    assign("rr_question_number", rr_question_number + 1, envir = .GlobalEnv)
    return(
      paste(
        "\n\n### ", 
        rr_homework_name[1], ", #", 
        rr_question_number, ": \"" , rr_question_chunk_name, "\"   ", 
        "{#", rr_question_chunk_name, "}", sep="" ))
    } else {
      ## code to be run after a chunk
      }
  })


```


## Preamble {#squawk}
In my opinion, one of the most useful binary operators
is `%in%`. However it is seldom discussed in detail in books about R
from the previous decade.  It's nice to see that Hadley has it front and 
center in the vocabulary section of his Advanced R book. 

It can be used to test if the elements of a vector A also occur (or do not occur)
in a vector B.  For example
```{r in-for-example, gather_answers = TRUE}
A <- c("a", "g", "Z", "y", "c")
B <- letters[1:5]
A %in% B
```


Here is a simple question.  Here is the setup:
```{r in-ex-1-set-up}
# imagine x look like this:
x <- c(1, 4, 18, 9)
# and y like this:
set.seed(10)
y <- sample(1:100, size = 30)
```

Now, here is the first question:
```{r in-ex-1-first, rr.question=TRUE}
x
y

my_answer <- submit_answer({
  # put your answer on the line between here
  x %in% y
  y %in% x
  y * x
  `+`(x,y)
  # and here
})
#print(chunker)
```


## Another part {#another-part}
```{r super-easy, rr.question=TRUE}
x <- 1:10
```


```{r}
save(boogiewho, file ="/tmp/boing.rda")
```
