---
title: "Homework 3"
author: "Eric C. Anderson"
output: bookdown::html_chapter
layout: default_with_disqus
---

```{r setup, echo=FALSE, include=FALSE}
library(knitr)
library(rrhw)
init_homework("Fun Homework 25")
rr_github_name <- NA
rr_pull_request_time <- NA
rr_question_chunk_name <- "NotSet"
```


## Preamble {#squawk}
In my opinion, one of the most useful binary operators
is `%in%`. However it is seldom discussed in detail in books about R
from the previous decade.  It's nice to see that Hadley has it front and 
center in the vocabulary section of his Advanced R book. 

It can be used to test if the elements of a vector A also occur (or do not occur)
in a vector B.  For example
```{r in-for-example, gather_answers = TRUE}
A <- c("a", "g", "Z", "y", "c")
B <- letters[1:5]
A %in% B

## and here I want to see if anything is happening:
rr_homework_name
rr_question_number
```


Here is a simple question.  Here is the setup:
```{r in-ex-1-set-up}
# imagine x look like this:
x <- c(1, 4, 18, 9)
# and y like this:
set.seed(10)
y <- sample(1:100, size = 30)
```

Now, here is the first question:
```{r in-ex-1-first, rr.question=TRUE}
x
y

# this should have gotten incremented:
rr_question_number

submit_answer({
  # put your answer on the line between here
  x %in% y
  y %in% x
  y * x
  `+`(x,y)
  # and here
})
#print(chunker)
```


## Another part {#another-part}
```{r super-easy, rr.question=TRUE}
x <- 1:10
```


```{r Boingus, rr.question=TRUE}
submit_answer({
  y<-1:10; z<- y^2; z;
  list(What = y, Never = "squawk", Again = z)
})
```


```{r Bogulated, rr.question=TRUE}
submit_answer({
  y<-1:10; z<- y^2; z;
  list(What = y, Never = "squawk", Again = z)
  xx <- 1:10; yy <- xx + rnorm(10, 0, .04)
  unclass(lm(yy~xx))
})
```




```{r}
ls(envir = rrhw_env)

# now make a list of it:
reslist <- as.list(rrhw_env)

library(plyr)
zz <- ldply(.data = reslist, .fun = answer_list_to_data_frame)
save(zz, file="/tmp/zz.rda")
```
